---
title: "Example Rmarkdown"
author: Sean Hardison
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---



```{r setup, include=FALSE}

# Use the `opts_chunk` function to set global parameters for code chunks. For instance, the following code results in all code chunks being printed into the rendered document along with output. If we wanted to ignore warning messages from code chunks, we could set `warning = FALSE`

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.align = 'center')

# load libraries
library(tidyverse)
library(plotly)
library(lubridate)
library(kableExtra)
library(DT)


# I prefer to load data and source scripts in the setup chunk. To do this, I make use of the `here` package in the following way:

#1. Identify the file path where data is stored. Using here() defaults the base directory to wherever your open .Rproj file is located. This means that any of us could download this repository and automatically be able to run this code without use of `setwd()`.
data.dir <- here::here("html_rmd","data")

#2. Load data
load(file.path(data.dir, "LME_nutrients_spatial.rdata"))
```

## Process data

If I can avoid it, I prefer to separate code chunks for processing and plotting. 

```{r data-processing}

#Process and summarise data-----------------------------------------------------------------
EcoMon_sum <- LME_nutrients_spatial %>% 
  filter(Value != -999, EPU %in% c("GB","GOM","SS")) %>% 
  
  #Create sample depth bins---------------------------------------------------------------------
  mutate(Value = as.numeric(Value),
         Depth_station = as.numeric(Depth_station),
         Depth_sampling = as.numeric(Depth_sampling)) %>% 
  mutate(bot_dif = Depth_station-Depth_sampling) %>% 
  mutate(surf_bot = ifelse(bot_dif <= 10, "Bottom",
                           ifelse(bot_dif > 10 & Depth_sampling <= 5, "Surface", "mid-water"))) %>%
  
  #Create seasonal bins-------------------------------------------------------------------------
  filter(month(Time) %in% c(10,11,5,6)) %>% 
  mutate(season = ifelse(month(Time) == 11 | month(Time) == 10, "Fall","Spring")) %>% 
  
  #Summarise------------------------------------------------------------------------------------
  group_by(EPU,
           season,
           surf_bot,
           Var,
           Year = year(Time)) %>% 
  dplyr::summarise(Value = mean(Value, na.rm = T)) %>% 
  as.data.frame()

```

## Plot data

Plots can be embedded easily.

```{r ecomon_data, fig.cap="Summarized nutrient and oceanographic data from the NEFSC EcoMon surveys in New Enlgand Ecological Production Units (EPUs).", out.width="90%",fig.height=8, fig.align='center'}
EcoMon_sum %>% 
  filter(!Var %in% c("BTLNBR","CASTNO","Cruise_ID","EXPOCODE","STNNBR")) %>% 
  ggplot() +
  geom_line(aes(x = Year, y = Value, color = surf_bot)) +
  guides(color = guide_legend("Sampling\n depth")) +
  facet_grid(Var~EPU + season, scales = "free") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.y = element_text(size = 6)) 
```

## Tables

So can tables!

```{r, echo = F}
vars <- data.frame(Variable = c("Cruise identifier","Cruise identifier","Station number",
                                "CTD cast number","Sample bottle number","Sample date",
                                "Sample time","Latitude","Longitude","Depth of station",
                                "Depth of sample","Water pressure","Water temperature",
                                "Water salinity","Potential density at surface pressure",
                                "Dissolved oxygen","Silicic acid concentration",
                                "Total nitrate and nitrite concentration","Ammonia concentration",
                                "Phosphate concentration","Dissolved oxygen"),
                   Names = c("EXPOCODE","Cruise_ID","STNNBR","CASTNO",
                             "BTLNBR","Date_UTC","Time_UTC",
                             "Latitude","Longitude","Depth_station",
                             "Depth_sampling","CTDPRS","CTDTEMP",
                             "CTDSAL","Sigma.Theta","CTDOXY",
                             "SILCAT","NITRIT+NITRAT","AMMMONIA",
                             "PHSPHT","CTDOXYMOL"),
                   Units = c("","","",
                             "","","MM/DD/YYYY",
                             "hh:mm","decimal degrees","decimal degrees",
                             "m","m","decibars","°C",
                             "PSS-78","kg m^-3^","mg L^-1^",
                             "$\\mu$M","$\\mu$M","$\\mu$M",
                             "$\\mu$M","$\\mu$mol kg^-1^"))
knitr::kable(vars,caption = "Variable definitions for the LME_nutrients_spatial.rdata file.")%>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  column_spec(2, width = "5cm") %>% 
  scroll_box(height = "400px")
```

## HTML Widgets

[HTML widgets](https://www.htmlwidgets.org/) use R functions as wrappers for JavaScript libraries. One of the benefits of knitting `rmarkdown` documents to HTML is that we're not limited to static images, and can take advantage of these widgets. The interactive graphing library `Plotly` is a good example of an html widget with huge flexibility for data visualization.

### Plotly 

```{r}
library(plotly)
# volcano is a numeric matrix that ships with R
p <- plot_ly(z = ~volcano) %>% add_surface()
p
```

### plotly::ggplotly()

Figures produced with `ggplot2` can be made interactive by passing `ggplot` objects to the `ggplotly` function. For example:

```{r}
n_gg <- EcoMon_sum %>% 
  filter(Var %in% c("NITRIT.NITRAT")) %>% 
  ggplot() +
  geom_line(aes(x = Year, y = Value, color = surf_bot)) +
  facet_grid(EPU~season)

ggplotly(n_gg) %>% 
  layout(legend = list(orientation = 'h',
                       y = -0.1))

```

### DT

[DT](https://rstudio.github.io/DT/) is another HTML widget that interfaces with the JS DataTables library. 

```{r}
library(DT)
datatable(vars)
```



## Tabbed output {.tabset}

Tabbed outputs can be added to save vertical space

### The Code

```{r, eval = T, echo = T}
cars_plot <- ggplot(data = mtcars) +
  geom_point(aes(x = wt, y = mpg)) +
  ggtitle("A plot")
```

### The Figure 

```{r, echo = F}
cars_plot
```