---
title: "tidyverse examples"
date: "3/24/2020"
output: html_document
---

Manipulate seabird data set using any tidyverse packages/functions.

 - [magrittr](https://www.rdocumentation.org/packages/magrittr/versions/1.0.0/topics/%25%3E%25)
 - [dplyr](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8)
 - [tidyr](https://tidyr.tidyverse.org/)

**1)** read in relevent packages, set up relative directory, read in csv
```{r, echo = T, eval = T, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyr)
library(dplyr)

raw.dir <- here::here("tidyverse")
seabird<- read.csv(file.path(raw.dir,"Audubon_Common_Tern.csv"))
head(seabird)
```

**2)** group_by() island and get productivity mean for each island
```{r, echo = T, eval = T, message=FALSE, warning=FALSE}
test.1<- seabird %>% 
  group_by(Island) %>% 
  summarise(mean =  mean(Productivity, na.rm = TRUE))

head(test.1)

```

**3)** filter for your fav island and then get the mean of that island's productivity
```{r, echo = T, eval = T, message=FALSE, warning=FALSE}
test.2<- seabird %>% 
  filter(Island == "EER") %>% 
  summarise(mean =  mean(Productivity, na.rm = TRUE))
head(test.2)

```

**4)** select relevent columns, mutate() = create new columns populated by the mean and SD productivity of each island
```{r, echo = T, eval = T, message=FALSE}
test.3<- seabird %>% 
  group_by(Island) %>% 
  mutate(mean = mean(Productivity, na.rm = TRUE),
         SD = sd(Productivity, na.rm = TRUE)) %>% 
  select(Island, Species, Productivity, mean, SD)
head(test.3)
```

**5)** pivot_longer = convert dataset to long format and filter out only diet species of interest 
```{r, echo = T, eval = T, message=FALSE, warning=FALSE}
test.4<- seabird %>% 
   tidyr::pivot_longer(cols = Productivity:Mackerel, names_to = "Var", values_to = "Value") %>% 
  filter(Var == c("Hake", "Pollock"))

head(test.4)
```

